---
title: "ST340 Lab 1: Time complexity"
date: "2019--20"
output: pdf_document
---
# 1: Implement bubblesort

`a` is a vector and the function should return `a` in increasing sorted order.
Example: if `a = c(3,5,2,4,1)`, then the output should be `c(1,2,3,4,5)`.

```{r}
bubble.sort <- function(a) {
  n <- length(a)
  if (n == 1) return(a)
  okay <- FALSE
  while (!okay) {
    okay <- TRUE
    
    for  ( i in 1:(n-1)) {
      if(a[i] > a[i+1]) {
        tmp <- a[i]
        a[i] <- a[i+1]
        a[i+1] <- tmp
        okay <- FALSE
      }
    }
  }
  return(a)
}
```

(a) Complete the function above.
(b) Test that it works.
    ```{r, eval=FALSE}
print(bubble.sort(c(3,5,2,4,1)))
print(bubble.sort(c(4,2,7,6,4)))
```
(c) Look at `?system.time`.
(d) How long does it take to sort `(1,2,...,10000)`?
(e) How about `(10000,1,2,3,...,9999)`?
(f) How about `(2,3,...,2000,1)`?
(g) How about a random permutation (see `?sample`) of 1,...,2000?
(h) Finally, recall the worst case input is `(n,n-1,...,2,1)`. Try the worst case input with `n = 2000`.

# 2: Implement quicksort

First, increase the maximum number of nested expressions that can be evaluated.
```{r}
options(expressions=100000)
```

`a` is a vector and the function should return `a` in increasing sorted order.
Example: if `a = c(3,5,2,4,1)`, then the output should be `c(1,2,3,4,5)`.

```{r}
qsort <- function(a) {
  if (length(a) > 1) {
    pivot <- a[1]
    # implement the pseudo-code in the lecture slides
  }
  return(a)
}
```

(a) Complete the function above.
(b) Test that it works.
    ```{r, eval=FALSE}
    print(qsort(c(3,5,2,4,1)))
    print(qsort(c(4,2,7,6,4)))
    ```
(c) <!--- Recall the worst case is when the input is sorted or reverse sorted-->
How long does it take to quicksort `(1,2,...,2000)`?
(d) How long does it take to quicksort `(2000,1999,...,1)`?
(e) How long does it take to quicksort a random permutation of `(1,2,...,2000)`?

# 3: Implement randomized quicksort

`a` is a vector and the function should return `a` in increasing sorted order.
Example: if `a = c(3,5,2,4,1)`, then the output should be `c(1,2,3,4,5)`.

```{r}
randomized.qsort <- function(a) {
  n <- length(a)
  if (n > 1) {
    pivot <- a[sample(n,size=1)]
    # implement the pseudo-code in the lecture slides
  }
  return(a)
}
```

(a) Complete the function above.
(b) Test that it works.
    ```{r,eval=FALSE}
    print(randomized.qsort(c(3,5,2,4,1)))
    print(randomized.qsort(c(4,2,7,6,4)))
    ```
(c) How long does it take to sort `(1,2,...,2000)`, `(2000,1999,...,1)`, or a random permutation, using randomized quicksort?

# 4: Compare the running time of the algorithms

Worst-case bubble and quicksort:

```{r}
ns <- seq(from=100,to=2000,by=100)
bubble.times <- rep(0,length(ns))
quick.times <- rep(0,length(ns))
randomized.quick.times <- rep(0,length(ns))
for (i in 1:length(ns)) {
  a <- ns[i]:1 # a is in reverse sorted order
  bubble.times[i] <- system.time(bubble.sort(a))[3]
  quick.times[i] <- system.time(qsort(a))[3]
  randomized.quick.times[i] <- system.time(randomized.qsort(a))[3]
}
```

(a) Plot `bubble.times` against `ns`, and also against `ns^2`.
(b) Plot `quick.times` against `ns`, and also against `ns^2`.
(c) Plot `randomized.quick.times` against `ns`.

# 5: Implement counting sort

`a` is a vector of positive integers and the function should return `a` in increasing sorted order.
Example: if `a = c(3,5,2,4,1)`, then the output should be `c(1,2,3,4,5)`.

```{r}
countingsort <- function(a) {
  n <- length(a); N <- max(a)
  # implement the pseudo-code in the lecture slides
  return(NULL)
}
```

# 6: Compare the running time of randomized quick sort and counting sort

```{r}
N <- 1e7 # maximum value of the positive integers
ns2 <- 1e5*(1:10)
randomized.quick.times2 <- rep(0,length(ns2))
counting.times2 <- rep(0,length(ns2))
for (i in 1:length(ns2)) {
  # each element of a is a draw from a categorical distribution
  a <- sample(N,size=ns2[i],replace=TRUE)
  counting.times2[i] <- system.time(countingsort(a))[3]
  randomized.quick.times2[i] <- system.time(randomized.qsort(a))[3]
}
```

(a) Plot `counting.times2` against `ns2`.
(b) Add `randomized.quick.times2` against `ns2` to the same plot.
(c) How would you describe the time complexity of randomized quick sort for the type of inputs generated above, assuming we only change `n`?
(d) Does this contradict the $\Omega(n\log n)$ lower bound discussed in class for comparison-based sorting algorithms?
