---
title: "ST340 Lab 5: Bandit problems"
date: "2019--20"
output: pdf_document
---
# Bernoulli Bandits

```{r}
library(mvtnorm)
```
(a) Set Bernoulli success parameters for each arm.
```{r}
ps <- c(0.4,0.6)
```

(b) This is a template for an Epsilon-greedy algorithm, runs for $n$ steps:
```{r}
epsilon.greedy <- function(ps,epsilon,n) {
  as <- rep(0,n)
  rs <- rep(0,n)
  
  ## initial number of plays and number of successes is 0 for each arm
  ns <- rep(0,2); ss <- rep(0,2)
  
  ## at first, play each arm once
  for (i in 1:2) {
    a <- i
    r <- runif(1) < ps[a]
    ns[a] <- ns[a] + 1
    ss[a] <- ss[a] + r
    as[i] <- a
    rs[i] <- r
  }
  
  ## now follow the epsilon greedy strategy
  for (i in 3:n) {
    # with probability epsilon, pick an arm uniformly at random
    if (runif(1) < epsilon) {
      a <- sample(2,1)
    } else { # otherwise, choose the "best arm so far".
      a <- which.max(ss/ns)
    }
    ## simulate the reward
    r <- runif(1) < ps[a]
    
    # update the number of plays, successes
    ns[a] <- ns[a] + 1
    ss[a] <- ss[a] + r
    
    # record the arm played and the reward received
    as[i] <- a
    rs[i] <- r
  }
  return(list(as=as,rs=rs))
}
```
Run `epsilon.greedy` with the given `ps` and a choice of `epsilon` and see how well it does.
```{r}
eg.out <- epsilon.greedy(ps=ps,epsilon=.1,n=1e4)
```
(c) Implement a `sample_arm` routine, for use in the Thompson sampling code below.
```{r}
sample_arm.bernoulli  <- function(ns,ss) {
  
  ## YOUR CODE HERE
  
}
```

``` {r}
thompson.bernoulli <- function(ps,n) {
  as <- rep(0,n)
  rs <- rep(0,n)
  
  ## number of times each arm has been played
  ## and number of corresponding successes
  ns <- rep(0,2); ss <- rep(0,2)
  
  for (i in 1:n) {
    a <- sample_arm.bernoulli(ns,ss)
    r <- runif(1) < ps[a]
    ns[a] <- ns[a] + 1
    ss[a] <- ss[a] + r
    as[i] <- a
    rs[i] <- r
  }
  return(list(as=as,rs=rs))
}
```

(d) Run the Thompson scheme and compare its performance to that of `epsilon.greedy`.
```{r,eval=FALSE}
thompson.bernoulli.out <- thompson.bernoulli(ps=ps,n=1e4)
```
