---
title: "Solutions for ST340 Lab 6"
date: "2019--20"
output: pdf_document
---
# Validation

The dataset `SmokeCancer.csv` shows lung cancer rates by U.S. state in 2010, with a number of covariates such as Federal Year 2010 cigarette sales per 100,000.

(a) Read the data file on lung cancer and create a data frame with variables of interest.
```{r}
X = read.table("SmokeCancer.csv", header=TRUE,sep=",",row.names=1)
LungCancer = data.frame(CigSalesRate=100000*X[,"FY2010Sales"]/X[,"Pop2010"],
                        X[,c("CigYouthRate","CigAdultRate","LungCancerRate")])
```

(b) Fit a linear model for LungCancerRate (`?lm` for a reminder about `lm`):
```{r}
summary(lm(LungCancerRate~CigSalesRate+CigYouthRate+CigAdultRate,data=LungCancer))
```

(c) Write a function that takes a formula and does LOOCV (leave one out cross validation) with respect to the squared error of the linear model for the given formula. Use it to find a good linear model for `LungCancerRate` in terms of `CigSalesRate`, `CigYouthRate` and `CigAdultRate`. You could also try using transformations of the covariates by adding terms such as `I(CigSalesRate^2)` and `I(CigSalesRate*CigAdultRate)` to your formulae. 

    (By good, we mean that it is the optimal, in terms of cross-validation error, linear model using some or all of these covariates.)

```{r}
loocv<-function(formula) {
  s=0
  for (i in 1:dim(LungCancer)[1]) {
    l=lm(formula,LungCancer[-i,])
    s=s+(predict(l,LungCancer[i,])-LungCancer$LungCancerRate[i])^2    
  }
  s/dim(LungCancer)[1]
}

loocv("LungCancerRate~CigSalesRate+             CigAdultRate")
loocv("LungCancerRate~CigSalesRate+CigAdultRate             +I(CigSalesRate^2)")
loocv("LungCancerRate~             CigAdultRate             ")
loocv("LungCancerRate~CigSalesRate+CigYouthRate+CigAdultRate")
loocv("LungCancerRate~CigSalesRate+CigAdultRate             +I(CigSalesRate*CigAdultRate)")
loocv("LungCancerRate~             CigYouthRate+CigAdultRate")
loocv("LungCancerRate~CigSalesRate+CigAdultRate             +I(CigAdultRate^2)")
loocv("LungCancerRate~CigSalesRate                          ")
loocv("LungCancerRate~CigSalesRate+CigYouthRate             ")
```

(d) The Akaike Information criterion (AIC) and Bayesian Information criterion (BIC) are analytic approximations to the validation step. They are (different) ways of quantifying the trade-off between model complexity (in terms of, e.g. the number of parameters) and the fit to the training data (in terms of likelihood), defined as follows:

* Akaike Information criterion (AIC) = $(2 \times \#\text{parameters} - 2 \times \log(\text{likelihood}))$, and 
* Bayesian information criterion (BIC) = $(\log(\text{amount of data}) \times \#\text{parameters} - 2 \times \log(\text{likelihood}))$

    Write a function that takes a formula and then calculates AIC and BIC. Use your function to find a *good* linear model for `LungCancerRate`, as in (b).

```{r}
aic<-function(formula) {
  AIC(lm(formula,data=LungCancer))
#   #Equivalent to 
#   l=lm(formula,data=LungCancer)
#   p=length(l$coefficients)+1
#   logLik(l)
#   2*p-2*logLik(l) # or:
#   2*p-2*sum(log(dnorm(l$residuals,sd=summary(l)$sigma)))
}

bic<-function(formula) {
  BIC(lm(formula,data=LungCancer))
}

aic("LungCancerRate~CigSalesRate+CigAdultRate             ")
aic("LungCancerRate~CigSalesRate+CigYouthRate+CigAdultRate")
aic("LungCancerRate~CigSalesRate+CigAdultRate             +I(CigAdultRate^2)")
aic("LungCancerRate~CigSalesRate+CigAdultRate             +I(CigSalesRate^2)")
aic("LungCancerRate~CigSalesRate+CigAdultRate             +I(CigSalesRate*CigAdultRate)")
aic("LungCancerRate~             CigAdultRate             ")
aic("LungCancerRate~             CigYouthRate+CigAdultRate")
aic("LungCancerRate~CigSalesRate                          ")
aic("LungCancerRate~CigSalesRate+CigYouthRate             ")

bic("LungCancerRate~CigSalesRate+             CigAdultRate")
bic("LungCancerRate~CigSalesRate+CigYouthRate+CigAdultRate")
bic("LungCancerRate~             CigYouthRate+CigAdultRate")
bic("LungCancerRate~             CigAdultRate             ")
bic("LungCancerRate~CigSalesRate                          ")
bic("LungCancerRate~CigSalesRate+CigYouthRate             ")
```

# The curse of dimensionality

Suppose $N$ points are chosen uniformly at random in the $D$-dimensional hypercube $[0,1]^D$. Consider a smaller hypercube $H = [0,r]^D$ in the "corner" of $[0,1]^D$.

(a) How big does $r$ have to be for there to be approximately one of the $N$ points lying in $H$?

> $(1/N)^{1/D}$.

(b) How big does $r$ have to be for there to be approximately 10 of the $N$ points lying in $H$?

> $(10/N)^{1/D}$.

(c) How big does $r$ have to be for there to be approximately $\frac{N}{2}$ of the $N$ points lying in $H$?

> $(1/2)^{1/D}$ which is approximately 1 for large $D$.

Check each of your answers by simulation.

```{r}
a1 = vector(); a2 = vector(); a3 = vector()
N = 10000
for (D in 1:10) {
  p = matrix(runif(N*D),nrow = N, ncol = D)
  r1 = (1/N)^(1/D)
  r2 = (10/N)^(1/D)
  r3 = (1/2)^(1/D)
  a1[D] = sum(rowSums(p < r1) == D) # Should average 1
  a2[D] = sum(rowSums(p < r2) == D) # Should average 10
  a3[D] = sum(rowSums(p < r3) == D) # Should average N/2
}
a1
a2
a3
```

# Distance functions

(a) Write a function to calculate the $\ell_1$ distances between pairs of row vectors in two matrices:
```{r eval=FALSE}
distances.l1 <- function(x,y) {
  apply(y,1,function(p) apply(x,1,function(q) sum(abs(p-q))))
}
```

(b) Write a similar function to calculate a matrix of pairwise $\ell_2$ distances:
```{r eval=FALSE}
distances.l2 <- function(x,y)
  apply(y,1,function(p) apply(x,1,function(q) sqrt(sum((p-q)^2))))
```

(c) Write a similar function to calculate the Mahalanobis distance between the row vectors, given a $D \times D$ covariance matrix $S$:
```{r eval=FALSE}
distances.maha <-function(x,y) {
  C=cov(x)
  C.inv=solve(C)
  apply(y,1,function(p) apply(x,1,function(q) sqrt( (p-q) %*% C.inv %*% (p-q) )))
}
```
